---
title: "Comparison of Raw and Corrected GC-IMS Data"
author: "Tecla Duran Fort"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    fig_caption: true
---

```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
knitr::opts_chunk$set(fig.path = "Comparison_corrected_files/figure-latex/")

# Load packages
library(ggplot2)
library(dplyr)
library(tidyr)
library(tibble)
library(reshape2)
library(ggpubr)

# Load GC-IMS tools (including correction() function)
source("../../load_gcims_tools.R")
```



## 1. Introduction

This report compares the **raw** and **corrected** GC-IMS Peak Table using the same evaluation metrics presented in the [Stability Analysis](https://github.com/tecladuran/gcims-workflows/blob/main/docs/stability_analysis.md), namely **Relative Standard Deviation (RSD)** and **variance explained by external factors**.

The correction applied is based on the orthogonalization procedure described in the [EPO Correction Report](https://github.com/tecladuran/gcims-workflows/blob/main/docs/epo_correction.md). Rather than repeating theoretical explanations, this document focuses on quantifying the improvement in signal stability and reduction of unwanted variability after correction.



## 2. Apply Correction

```{r apply-correction}
df <- read.csv("../../data/peak_table_var.csv")
intensities <- df %>% select(starts_with("Cluster"))

# Apply sequential correction
corr_time <- correction(intensities, df$elapsed_time)
intensities_time_corr <- corr_time$corrected

corr_batch <- correction(intensities_time_corr, df$batch)
intensities_final <- corr_batch$corrected
```

---

## 3. Relative Standard Deviation (RSD)

```{r rsd-comparison, echo=FALSE}
rsd_raw <- sapply(intensities, function(x) sd(x) / mean(x) * 100)
rsd_corr <- sapply(intensities_final, function(x) sd(x) / mean(x) * 100)

rsd_df <- data.frame(Cluster = names(rsd_raw),
                     Raw = rsd_raw,
                     Corrected = rsd_corr)

rsd_melt <- rsd_df %>% pivot_longer(-Cluster, names_to = "Condition", values_to = "RSD")

ggplot(rsd_melt, aes(x = RSD, fill = Condition)) +
  geom_histogram(position = "identity", binwidth = 5, alpha = 0.6) +
  scale_fill_manual(values = c("#1D3557", "#A8DADC")) +
  labs(title = "RSD Distribution Before and After Correction",
       x = "Relative Standard Deviation (%)",
       y = "Number of Clusters") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(hjust = 0.5, size = 12))
```

---



## 4. Cluster Stability 

To evaluate how much the correction improves the technical robustness of the dataset, we compute the number of clusters whose Relative Standard Deviation (RSD) falls below the 20% threshold, both **before** and **after** correction. This threshold is commonly used as an orientative benchmark in metabolomics quality control.

```{r rsd-summary-table, echo=FALSE, fig.align="center", fig.width=5, fig.height=1, fig.cap = "Proportion of clusters with RSD below 20%, before and after correction"}

# Compute number of stable clusters
threshold <- 20
raw_stable <- sum(rsd_raw < threshold)
corr_stable <- sum(rsd_corr < threshold)
total <- length(rsd_raw)

stable_df <- data.frame(
  Condition = c("Raw Data", "Corrected Data"),
  Stable_Clusters = c(raw_stable, corr_stable),
  Proportion = paste0(round(c(raw_stable, corr_stable) / total * 100, 1), " %")
)

ggtexttable(stable_df, rows = NULL, theme = ttheme("light"))
```

The correction process increases the proportion of stable clusters from **`r round(raw_stable / total * 100, 1)`%** to **`r round(corr_stable / total * 100, 1)`%**, confirming that the removal of variance associated with acquisition order and batch improves overall signal reliability.

---

## 5. Explained Variance by External Factors

```{r explained-variance, echo=FALSE}
explained_variance <- function(data, variable){
  apply(data, 2, function(x){ summary(lm(x ~ variable))$r.squared }) %>% mean()
}

variance_df <- data.frame(
  Condition = c("Raw", "Corrected"),
  Elapsed_Time = c(explained_variance(intensities, df$elapsed_time),
                   explained_variance(intensities_final, df$elapsed_time)),
  Batch = c(explained_variance(intensities, df$batch),
            explained_variance(intensities_final, df$batch))
)

variance_df[,2:3] <- round(variance_df[,2:3] * 100, 2)

ggtexttable(variance_df, rows = NULL, theme = ttheme("light"))
```

---

## 6. Principal Component Analysis (PCA)

We now perform a new PCA on the corrected data to explore whether the dominant sources of variation are still aligned with external variables. The PCA on raw data already showed strong trends related to `elapsed_time` and `batch`, as shown in previous reports.

### 6.1 PCA of Raw Data 
```{r pca-raw, echo=FALSE, warning=FALSE}
# Perform PCA on raw data

pca_raw <- prcomp(scale(intensities))
explained_var_raw <- summary(pca_raw)$importance[2, ] * 100
axis_labels <- paste0("PC", 1:2, " (", round(explained_var_raw[1:2], 1), "%)")

pca_df_raw <- data.frame(
  PC1 = pca_raw$x[,1],
  PC2 = pca_raw$x[,2],
  Elapsed_Time = df$elapsed_time,
  Batch = as.factor(df$batch)
)
```

```{r pca-raw-plots, echo=FALSE, fig.width=12, fig.height=4.5, fig.align='center', fig.cap = "PCA of raw data colored by elapsed time (left) and by batch (right)"}
p1 <- plot_pca(pca_df_raw, "Elapsed_Time", "Colored by Elapsed Time",
               scale_color_gradientn(colours = hcl.colors(100, "YlGnBu")))

p2 <- plot_pca(pca_df_raw, "Batch", "Colored by Batch",
               scale_color_discrete())

p1 + p2
```

```{r pca-variance-raw, echo=FALSE, fig.width=5, fig.height=3, fig.align='center', fig.cap = "Explained variance by the first six PCA components (raw data)"}
var_df_raw <- data.frame(
  PC = paste0("PC", 1:6),
  Variance = explained_var_raw[1:6]
)

ggplot(var_df_raw, aes(x = PC, y = Variance)) +
  geom_col(fill = "#A8DADC") +
  theme_minimal(base_size = 13) +
  labs(title = "Explained Variance — Raw PCA",
       x = "Principal Component", y = "Explained Variance (%)") +
  theme(axis.text.x = element_text(size = 11))
```

---

### 6.2. PCA of Corrected Data
```{r pca-new, echo=FALSE}
# PCA on corrected data
pca_corr <- prcomp(scale(intensities_final))
explained_corr <- summary(pca_corr)$importance[2, ] * 100
axis_labels_corr <- paste0("PC", 1:2, " (", round(explained_corr[1:2], 1), "%)")

pca_df_corr <- data.frame(
  PC1 = pca_corr$x[, 1],
  PC2 = pca_corr$x[, 2],
  Elapsed_Time = df$elapsed_time,
  Batch = as.factor(df$batch)
)
```


```{r pca-corrected-color, echo=FALSE, fig.width=12, fig.height=4.5, fig.cap = "PCA of corrected data colored by elapsed time (left) and by batch (right)", fig.align='center'}

p1 <- plot_pca(pca_df_corr, "Elapsed_Time", "Colored by Elapsed Time",
               scale_color_gradientn(colours = hcl.colors(100, "YlGnBu")))

p2 <- plot_pca(pca_df_corr, "Batch", "Colored by Batch",
               scale_color_discrete())

p1 + p2
```

```{r pca-corrected-variance, echo=FALSE, fig.width=5, fig.height=3, fig.cap = "Variance explained by the first six PCA components (corrected data)", fig.align='center'}

var_df_corr <- data.frame(PC = paste0("PC", 1:6),
                          Variance = explained_corr[1:6])

ggplot(var_df_corr, aes(x = PC, y = Variance)) +
  geom_col(fill = "#A8DADC") +
  theme_minimal(base_size = 13) +
  labs(title = "Corrected PCA — Variance Explained",
       x = "Principal Component", y = "Explained Variance (%)") +
  theme(axis.text.x = element_text(size = 11))
```

Compared to the PCA of the raw data, the corrected data shows a more homogeneous distribution of variance across components, and no evident separation or gradient is observed when coloring by elapsed time or batch. This suggests that external influences no longer dominate the variance structure after correction.

---


## 7. Conclusion

The applied correction successfully reduces the influence of time and batch effects. The PCA projection confirms that corrected data no longer follows the acquisition order. This confirms the effectiveness of the orthogonalization strategy when applied directly to the raw intensity matrix.


